---
title: 'Lab 2: exploring the US elections'
author: "(Your name)"
format: docx
execute:
  warning: false
editor_options: 
  chunk_output_type: console
---


*Note: The chunks below are not complete, you will need to fill in the blanks and knit to an output document to finish the lab!*

# Exploring the primaries

In this lab we will explore the 2020 US presidentials elections.
The goal is to see whether Biden was more successful in larger states. 

## Getting data

First, let's download the data from [MIT election lab](https://electionlab.mit.edu/data)'s [Dataverse page](https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/42MVDX).
Note that the dataverse requires you to sign their guest book before downloading, 
I make a copy that can be downloaded directly 
(which is allowed since it's CC-0):

```{r data}
library(tidyverse)
elections <- read_csv("https://raw.githubusercontent.com/vanatteveldt/auc_labs/refs/heads/main/data/1976-2024-president.csv")
head(elections)
```

As you can see, this has a lot of columns that we don't need. 
Use `select` to select only the `year`, `state`, `state_po`, `party_simplified`, `candidatevotes`, `totalvotes` columns.
Rename `party_simplified` to `party`

```{r clean}
# Insert your code here
elections = ...
```

Now, filter the data for the year 2020 and the democratic party:

```{r filter}
# Insert your code here
```

To check, display (the top rows of) the data set:

```{r check}
head(elections)
```

## Computing values

Now, use `mutate` to compute the percentage of total votes that Biden (the democratic candidate) received:

```{r compute totals}
# Insert your code here
```

(As a bonus, you can use `str_to_title` to normalize the capitalization in the party and candidate names)

## Exploring total results

Using `arrange` and `head` and/or `tail`, show the states where Biden had the largest and smallest vote share (percentage):

```{r exploration}
# Insert your code here
```

[your interpretation]

## Statistical analysis

Now, using `cor.test`, see if there is a correlation between total number of votes and Bidens's vote share. 
Note that `cor.test` is not part of the tidyverse, so it uses the `dataset$column` notation to select the two columns. 

Interpret the result briefly (1 or 2 sentences). What does `totalvotes` actually proxy for in this analysis?

```{r correlation}
# Insert your code here
```

[your interpretation]

## Bonus

As a bonus, can you recreate the graph from the assignment page on canvas? 
If you use AI, make sure that you understand what the code you get actually does!
If needed, ask AI what each line does, why it made certain choices, and play with the code yourself. 
