---
title: 'Lab 3: Visualizing US politics'
author: "(Your name)"
format: docx
execute:
  warning: false
editor_options: 
  chunk_output_type: console
---

# Introduction

The goal of this lab is to demonstrate your understanding of basic ggplot plots. 
For the first plots, don't worry about layout, themes, labels etc. -- just create the right graph. 
This is generally what you do during your own analysis/exploration phase.
In the last part you are asked to make one graph suitable for publication/presentation,
i.e. with a nice theme, good labels, good scales, etc.

# Data: US per-county Presidential election results

# Line graph: Changes over time

Changes over time are most often depicted as a line graph, e.g. using `geom_line()`. 
Consider this data set of county-level US election results over time,
obtained again from the [MIT election lab](https://electionlab.mit.edu/data): 

```{r}
library(tidyverse)
results_per_county <- read_csv("https://raw.githubusercontent.com/vanatteveldt/r_labs/main/data/countypres_2000-2024_clean.csv")
```

Compute a new data set, 'results_per_year', that computes the overall vote share per party per year.
Hint: you can first summarize the total candidatevotes grouped by party x year, and then divide by total candidatevotes per year. 

```{r}
# Add your data processing code here:
# results_per_year <- ...
```

Create a line graph to show the result of the parties over time, giving each party it's own line (in a different color). 

```{r}
# insert your code here
```

Bonus challenges:

- Can you change the color so the republicans (red) and democrats (blue) are shown in their traditional colors?
- Do all data years look OK? Is 2024 plausible? Can you find out what's wrong by filtering and inspecting the original and/or aggregated data set?
- If you look at the data, you can see that the smaller parties don't actually have data for each year. By default, ggplot just draws a line between the known points, but in this instance we should probably assume that party had zero votes. Can you use the `complete` function to make sure they are displayed as zeroes?

# Bar plots

Using the `results_per_county` data above, 
create a bar plot of how many county each party won in 2024.

(Hint: first filter on year, then compute the winner for each county using `group_by` and `slice_max`; then use `geom_col` to create the bar plot)

```{r}
# insert your code here
```

Bonus challenges: 
- Can you color by party, make the parties properly capitalized, and remove the axis labels and add a descriptive title?
- Substantively, where did the smaller parties go? Are there NA results? Why?
- Is there a difference between grouping by county_name and county_fips? Why?

# Scatter plot: Relation between variables

Relationships between unordered numerical variables are often shown in a scatter plot. 

## Data

The following code creates a combined data set of Trump vote share and selected demographics per county. 
This data is obtained from the [Census Bureau Data API](https://api.census.gov) but this usage is not endorsed or certified by the Census Bureau.

Note: we will discuss the `inner_join` and other join commands next week, but you can peak ahead in [R4CSS chapter 7](https://r4css.vanatteveldt.com/chapters/07-reshaping.html#joining-multiple-data-sets-using-_join)) if you are curious!)

```{r}
# (note: you don't need to change any code here since we didn't discuss joins yet)
demographics <- read_csv("https://raw.githubusercontent.com/vanatteveldt/r_labs/main/data/US_Demographics.csv")

trump_predictors <- results_per_county |>
  filter(party == "REPUBLICAN", year == 2024) |>
  group_by(county_fips) |> 
  mutate(trump_voteshare=candidatevotes/totalvotes,
         fips=sprintf("%05d", county_fips)) |>
  ungroup() |>
  select(fips, trump_voteshare) |>
  left_join(demographics, join_by(fips)) 
head(trump_predictors)
```

Bonus challenges: 
- Did the `left_join` drop any counties? Why?
- What does the sprintf command do, and why was it needed?

## Explaning Trump's vote share

Choose at least two variables that you think might predict or correlate with Trump's vote share. 
Create a scatter plot (using `geom_point`) to show this relation, with the vote share on the Y axis, one of the demographics on the x-axis, and the other demographics using a different aesthetic (e.g. `color` or `size`). 

```{r}
# insert your code here
```

# Make it beautiful

Finally, take the last plot and make it suitable for publication. 
This means that it should be pleasing to look at, with no badly chosen colors or scales and no technical terms or distracting artefacts. 
Moreover, it means that it should be self-explanatory: any interested reader should be able to read the story from just looking at the graph. 

This generally means at least adding an overall title (describing the main relation expressed in the graph), decent labels for the x and y axes, and decent names for other variables or legends. You could also consider changing the axis display, adding a subtitle, changing the color scheme, etc. Finally, feel free to apply or customize any theme that you think makes it look better.

```{r}
# insert your code here
```
