---
title: 'Lab 2: exploring the US elections'
author: "(Your name)"
format: docx
execute:
  warning: false
editor_options: 
  chunk_output_type: console
---


*Note: The chunks below are not complete, you will need to fill in the blanks and knit to an output document to finish the lab!*

# Exploring the primaries

In this lab we will explore the 2020 US presidentials elections.
The goal is to see whether Biden was more successful in larger states. 

## Getting data

First, let's download the data from [MIT election lab](https://electionlab.mit.edu/data)'s [Dataverse page](https://dataverse.harvard.edu/dataset.xhtml?persistentId=doi:10.7910/DVN/42MVDX):

```{r data}
# install.packages("dataverse")   # if needed)
library(dataverse)
elections = get_dataframe_by_name(
  filename = "1976-2020-president.tab",
  dataset = "doi:10.7910/DVN/42MVDX", 
  server = "dataverse.harvard.edu")
elections
```

Use `select` to select only the `year`, `state`, `state_po`, `party_simplified`, `candidatevotes`, `totalvotes` columns.
Rename `party_simplified` to `party`

```{r clean}
# Insert your code here
elections = ...
```

Now, filter the data for the year 2020 and the democratic party:

```{r filter}
# Insert your code here
```


To check, display (the top rows of) the data set:

```{r check}
head(elections)
```

## Computing values

Now, use `mutate` to compute the percentage of total votes that Biden (the democratic candidate) received:

```{r compute totals}
# Insert your code here
```

## Exploring total results

Using `arrange` and `head` and/or `tail`, show the states where Biden had the largest and smallest vote share (percentage):

```{r exploration}
# Insert your code here
```

[your interpretation]

## Statistical analysis

Now, using `cor.test`, see if there is a correlation between total number of votes and Bidens's vote share. Interpret the result briefly (1 or 2 sentences)

```{r correlation}
# Insert your code here
```

[your interpretation]

