---
title: "2026 AUC Lab 5: Working with texts"
format: html
editor_options: 
  chunk_output_type: console
---

As Wolfsfeld would put it: 
The conflict over the media is an essential part of politics.
So, being able to analyze media **content** is an important part of analysing political communication.

In this lab, you will use the `stringr` package to analyse social media messages,
*in casu* Tweets made by Trump when he was still on the platform. 
See [chapter 9 of R4CSS](https://r4css.vanatteveldt.com/chapters/09-stringr.html) 
and/or the [StringR cheat sheet](https://rstudio.github.io/cheatsheets/strings.pdf) for more information.

## Data

Although tweets are normally not allowed to be distributed,
the [Trump Archive](https://www.thetrumparchive.com/) makes the full version of all Trump's tweets 
available as a simple direct download.

```{r}
library(tidyverse)
tweets <- read_csv("https://drive.google.com/uc?export=download&id=1xRKHaP-QwACMydlDnyFPEaFdtskJuBa6")
head(tweets)
```

## Detecting Nicknames

Trump was famous for using nicknames like Sleepy Joe and Crooked Hillary for his opponents. 

First, let's make a cleaned version of the data set by filtering out all posts he retweeted (`isRetweet == FALSE` or just `!isRetweet`),
converting text to lower case with `str_to_lower` and normalizing whitespace with `str_squish`.  Store this as a separate data frame. 

Then:
1. Using `str_detect`, create a column for each nickname, 1 if mentioned, 0 if not (hint: as.numeric turns `TRUE`/`FALSE` into `1`/`0`).
2. Using `floor_date`, create a column for the month.
3. Using `summarize`, aggregate the sum of each nickname per month.
4. Using `pivot_longer`, turn this into a long data frame.
5. Using `ggplot`, create a line graph of each nickname over time.

```{r}
```

## Nicknames and engagement

Does the use of nicknames produce higher engagement? 
Using the cleaned dataset from above:

1. Use `str_extract` to extract the first nickname used to a new variable (hint: `cat|dog` extracts cat OR dog)
2. Filter to keep only tweets with a nickname
3. Calculate engagement as a sum of `favorites` and `retweets`
4. Make a scatter plot of individual tweets, showing engagement on the y axis, time on the x axis, and use color for the use of a nickname. 

For bonus points:
- Add a layer of grey dots below the nickname dots representing tweets without a nickname
- Place engagement on a logarithmic axis


```{r}
```

Do the results make sense? Can you check one or two 'sleepy' tweets from before 2015 -- were they actually about Joe Biden?

Using `str_extract`, what happens if a tweet has multiple nicknames? How often does that happen? How could you fix that?

## Using a simple pattern to extract @mentions

Trump is often accused of applying the label `fake news` to news organizations. Starting again from the cleaned data:

1. Filter to only keep tweets using fake news (or optionally variations of the term)
2. Using `str_extract` with a simple regular expression pattern, extract any at-mentions in these tweets
3. Make a table of how often each at-mention occurs, showing the top-10.

```{r}
```

Do the results make sense? In the posts mentioning Fox, are they actually called fake, or is it contrasting them?

For bonus points: 
- use `str_extract_all` and `unnest_longer` instead of `str_extract`. How does it change your results?
- contrast the mention within 'fake news' tweets to the overall mentions

## Making a word plot

Word clouds are overused and (deservedly) much maligned. But let's make a word plot anyway!

Starting again from the cleaned data:

1. Extract all hash tags using `str_extract_all` and `unnest_longer`.
2. For each hash tag, compute overall use and median date of use.
3. Using the `geom_text_repel` from the `ggrepel` package, plot the top-50 tags
   Use a random y axis, the date on the x axis, and the frequency as the size.

```{r}
```

## Sentiment analysis

Trump is often said to tweet with many negative emotions,
and interestingly this is [said to vary by device](http://varianceexplained.org/r/trump-tweets/): 
He personally types tweets on his Android, while his aides use an iPhone.

(Note that the article uses the old twitter API, which is now defunct;
it also uses `tidytext`, which we will introduce after the break. 
So, use it for inspiration more than for the code)

Can we use stringr to do a simple sentiment analysis per device? Starting again from the cleaned data:

1. Make a simple pattern to detect anger (or other negative emotions). Don't overthink it: even just `stupid|angry` is fine;
   you can use the NRC lexicon mentioned in the article, 
   or one of the many existing lists of negative emotions you can find online.
2. Filter out all results after March 2017.
3. Count in what percentage of tweets an angry emotion is present, grouped by device.
4. Make a barplot of the result.

```{r}
```

Do the results make sense? 
Do you think the sentiment analysis is valid?
How could you improve it?
