---
title: "AUC Lab1"
author: "Your names here"
format: docx
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning=FALSE)
library(printr)
```


# Introduction

This assignment consists of two parts, with two learning goals.

First, we ask you to run a very simple analysis on a data set of the last Dutch elections. 
The main goal is to show that you have installed R, can work with Quarto documents, 
and can run simple commands. 

Next, we ask you to look at a quite complex piece of R code. 
You are not expected to (completely) understand this code. 
The goal is to use Claude (or another genAI tool of your choice)
to answer a number of questions about the code,
and then reflect on your use of AI. 

## Part 1: 

### Install RStudio and Tidyverse

Install RStudio and Tidyverse on your computer. You can follow [these instructions](https://rstudio-education.github.io/hopr/starting.html) or any other instructions
to install R and RStudio.

To install tidyverse, you can use the 'Packages' pane in RStudio,
or run `install.packages("tidyverse")`. 
Do *not* include this command in your quarto submission!

If you already have R and tidyverse installed, you might want to update it to ensure you have a recent version.

### Data

We will use a simplified version of the 'Dutch elections' data from the R4CSS book:

```{r data}
library(tidyverse)
votes <- read_csv("https://raw.githubusercontent.com/vanatteveldt/auc_labs/refs/heads/main/data/dutch_votes_combined.csv")
head(votes)
```

This shows the vote share of each party in each municipality, as well as the total population and the percentage Dutch-born population in the town.

### Filtering and sorting

Choose a political party. 
Edit the code below to **filter** the data set to show only results from that party,
and **arrange** the data set by descending votes:

```{r party}
votes_party <- filter(votes)
votes_party <- arrange(votes_party)
head(votes_party)
```

Where does that party do well? Look up the party and one or two of the municipalities on wikipedia:
does it make sense that that party would do well there? 

**Enter your findings and explanation here (2-3 sentences)**

### Scatter plot

Now, let's see to what extent the party is more successful in larger or smaller municipalities,
*or* (your choice) in muncipalities with a higher or lower share of Dutch born citizens. 

First, let's first create a scatter plot to see to what extent their voting seems related.
In the code below, remove the `#` sign and fill in the blanks so the population is on the x-axis, 
and either the vote share or percentage Dutch are on the y-axis:

```{r scatter}
# ggplot(votes_party, aes(x=___, y=___)) + geom_point()
```

Interpretation: How do you interpret the scatter plot? Does there seem to be a relation between population size and votes?

**Enter your findings and explanation here (2-3 sentences)**

Bonus: If you want, you can add `+ labs(title="my title", y="Vote share", x="Population") + theme_classic()` to the plot to add labels and change the theme. Feel free to make other changes to the plot as you see fit!

## Part 2: Understanding R with Claude

The code below is from the visualization exercise at the end of Chapter 8 of R4CSS.
Substantively, it is a reproduction of Figure 1 from a 538 article on the Bechdel test:
[The Dollar-And-Cents Case Against Hollywood's Exclusion of Women](https://archive.ph/e7xDQ).
It shows how many films from each half-decade pass the 'Bechdel test':
Is there any scene in which two women talk to each other about something else than a man?

Run the code below and have a look at the result:

```{r}
bechdel <- read_csv("https://raw.githubusercontent.com/vanatteveldt/r4css/refs/heads/main/data/bechdel.csv")

fig1data <- bechdel |>
  mutate(test = str_remove(test, "-disagree"),
         lustrum = floor(year / 5) * 5) |>
  group_by(lustrum) |>
  mutate(start = min(year),
         end = max(year),
         period = str_glue("{start}-\n'{str_sub(end, 3, 4)}")) 

colors <- c(nowomen = "#ff2700", notalk = "#ff9380", men = "#ffc9bf",
            dubious = "#6bb2d5", ok = "#008fd5")
periods <- fig1data |> pull(period) |> unique() |> sort()
breaks <- periods[seq(1, length(periods), by = 2)]

ggplot(fig1data) +
  geom_bar(aes(x = period, fill = factor(test, levels = names(colors))), position = "fill") +
  scale_fill_manual(values = colors, guide = "none") +
  labs(title = "The Bechdel Test Over Time",
       subtitle = "How women are represented in movies") +
  scale_y_continuous(name = "", labels = scales::label_percent()) +
  scale_x_discrete(name = "", breaks = breaks) +
  annotate("text", label = "FAIL", y = .75, x = 5, size = 20, fontface = "bold") +
  annotate("text", label = "PASS", y = .25, x = 5, size = 20, fontface = "bold") +
  theme_minimal() +
  theme(panel.grid = element_blank())
```

Now, use [claude.ai](claude.ai) or a favorite genAI tool of your choice to answer the following questions.
Make sure to use an online chat tool (not claude code) 
and start a new session / chat with the tool as you will have to share the log at the end.
Also make sure not to put anything personal (or incriminating :D) in the chat!

1. What does aes do?
2. how are data, geoms, and aesthetics related in ggplot?
3. What does `fill = factor(test, levels = names(colors))` do? 
4. What is a 'factor', and why are the test outcomes converted to a factor here?
5. What does `scale_fill_manual` do? 
6. How do scales relate to data, geoms and aesthetics?
6. What is the difference between an `annotate` and a `geom`?
7. What would you need to change to make fail red, and pass blue?

For each question, don't copy paste the AIs answer: 
write it in your own words, and make sure you fully understand the answer.
You can ask additional clarification questions as needed. 
Limit your answer to one or two sentences for each question. 

### Chat log

Paste a link to your chat log here: ...

(use the 'share' icon on the conversation in Claude or your AI tool)

### Reflection (1 paragraph)

How useful was the AI in understanding complex code?
What questions were most useful in understanding the R code? 
How can you use AI in a good way to help learn R?

## Submitting this assignment

To submit the assignment, **Render** it to a word document, download it, and submit it on canvas. 
Alternatively, you can print the html report as a pdf and submit that, or install latex on your computer and knit directly to pdf. 
